Modal type theory based on the intuitionistic epistemic logic

### 2018-01-16 article.tex
finished subject reduction
added some beauty
added eta rules
extended proof
added interpretation for typing rules

  \begin{center}
  \begin{prooftree}
  \AxiomC{$ $}
  \RightLabel{\scriptsize{ax}}
  \UnaryInfC{$\Gamma , x : \alpha \vdash x : \alpha$}
  \end{prooftree}
  \end{center}

  \begin{minipage}{0.5\textwidth}
    \begin{flushleft}
  \begin{prooftree}
  \AxiomC{$\Gamma, x : \alpha \vdash M : \beta$}
  \RightLabel{$\rightarrow_i$}
  \UnaryInfC{$\Gamma \vdash \lambda x. M : \alpha \to \beta$}
  \end{prooftree}

  \begin{prooftree}
  \AxiomC{ $\Gamma \vdash x : \alpha$ }
  \AxiomC{ $\Gamma \vdash y : \beta$ }
  \RightLabel{$\times_i$}
  \BinaryInfC{$\Gamma \vdash (x, y) : \alpha \times \beta$}
  \end{prooftree}

  \begin{prooftree}
    \AxiomC{$\Gamma \vdash x : \alpha$}
    \RightLabel{$\textbf{K}_I$}
    \UnaryInfC{$\Gamma \vdash {\bf pure \: } \: x : \textbf{K}\alpha $}
  \end{prooftree}
    \end{flushleft}
  \end{minipage}

  \begin{minipage}{0.5\textwidth}
    \begin{flushright}
  \begin{prooftree}
  \AxiomC{$\Gamma \vdash f : \alpha \to \beta$}
  \AxiomC{$\Gamma \vdash x : \alpha$}
  \RightLabel{$\rightarrow_e$}
  \BinaryInfC{$\Gamma \vdash fx : \beta$}
  \end{prooftree}

  \begin{prooftree}
  \AxiomC{ $\Gamma \vdash p : \alpha_1 \times \alpha_2$ }
  \RightLabel{$\times_e$, $i \in \{ 1, 2 \}$}
  \UnaryInfC{$\Gamma \vdash \pi_i p : \alpha_i$}
  \end{prooftree}

  \begin{prooftree}
    \AxiomC{$\Gamma \vdash \vec{M} : {\bf K} \vec{A}$}
    \AxiomC{$\vec{x} : \vec{A} \vdash M : B$}
    \RightLabel{$\text{let}_{\bf K}$}
    \BinaryInfC{$\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \: } M : {\bf K} B$}
  \end{prooftree}
    \end{flushright}
  \end{minipage}
  \end{defin}

@@

  10) ${\bf let \: pure \:} x = M {\: \bf in \: } x \rightarrow_{\eta} M$;

  11) ${\bf pure \:} (\lambda x. f x) \rightarrow_{\eta} {\bf pure \:} f$;

  12) ${\bf pure \:} (\langle \pi_1 P, \pi_2 P \rangle) \rightarrow_{\eta} {\bf pure \:} P$;

  13) ${\bf pure \:} ({\bf let \: pure \:} x = M {\: \bf in \: } x) \rightarrow_{\eta} {\bf pure \:} M$.
  \end{defin}

@@

  1) Let $\Gamma \vdash x : A$, where $\Gamma = \Delta, x : \alpha$, $x \in \mathbb{V}$.

@@

  Let $\Gamma \vdash M : \alpha$ and $M \twoheadrightarrow_{\beta} N$, then $\Gamma \vdash N : \alpha$

@@

1) Let $\Gamma \vdash {\bf let \: pure \:} \langle \vec{x}, y, \vec{z} \rangle = \langle \vec{M}, {\bf let \: pure \: } \vec{w} = \vec{N} {\bf \: in \: } Q, \vec{P} \rangle {\: in \:} R : {\bf K }B$

By generation we have $\Gamma \vdash \vec{M} : {\bf K}\vec{A_1}$, $\Gamma \vdash {\bf let \: pure \: } \vec{w} = \vec{N} {\bf \: in \: } Q : {\bf K}\vec{A_2}$, $\Gamma \vdash \vec{P} : {\bf K}\vec{A_3}$
and $\vec{x} : \vec{A_1}, y : A_2, \vec{z} : \vec{A_3} \vdash R : B$.

@@

Hence $\Gamma \vdash {\bf pure \: } M_i : { \bf K } A_i$ by ${\bf K}_I$
  \end{proof}

@@

>>>>

    \begin{center}
    \begin{prooftree}
    \AxiomC{$ $}
    \RightLabel{\scriptsize{ax}}
    \UnaryInfC{$\Gamma , x : A \vdash x : A$}
    \end{prooftree}
    \end{center}

    \begin{minipage}{0.45\textwidth}
      \begin{prooftree}
      \AxiomC{$\Gamma, x : A \vdash M : B$}
      \RightLabel{$\rightarrow_i$}
      \UnaryInfC{$\Gamma \vdash \lambda x. M : A \to B$}
      \end{prooftree}

      \begin{prooftree}
      \AxiomC{ $\Gamma \vdash M : A$ }
      \AxiomC{ $\Gamma \vdash N : B$ }
      \RightLabel{$\times_i$}
      \BinaryInfC{$\Gamma \vdash \langle x, y \rangle : A \times B$}
      \end{prooftree}

      \begin{prooftree}
        \AxiomC{$\Gamma \vdash x : A$}
        \RightLabel{$\textbf{K}_I$}
        \UnaryInfC{$\Gamma \vdash {\bf pure \: } \: x : {\bf K}A $}
      \end{prooftree}
  \end{minipage}%
  \hfill
  \begin{minipage}{0.45\textwidth}
  \begin{tabular}{p{\textwidth}}
    \begin{prooftree}
    \AxiomC{$\Gamma \vdash f : A \to B$}
    \AxiomC{$\Gamma \vdash x : A$}
    \RightLabel{$\rightarrow_e$}
    \BinaryInfC{$\Gamma \vdash fx : B$}
    \end{prooftree}

    \begin{prooftree}
    \AxiomC{ $\Gamma \vdash M : A_1 \times A_2$ }
    \RightLabel{$\times_e$, $i \in \{ 1, 2 \}$}
    \UnaryInfC{$\Gamma \vdash \pi_i M : A_i$}
    \end{prooftree}

    \begin{prooftree}
      \AxiomC{$\Gamma \vdash \vec{M} : {\bf K} \vec{A}$}
      \AxiomC{$\vec{x} : \vec{A} \vdash M : B$}
      \RightLabel{$\text{let}_{\bf K}$}
      \BinaryInfC{$\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \: } M : {\bf K} B$}
    \end{prooftree}
  \end{tabular}
  \end{minipage}%

@@

  10) ${\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} N \rightarrow_{\eta} {\bf pure \:} N$;

  11) ${\bf let \: pure \:} x = M {\: \bf in \: } x \rightarrow_{\eta} M$;

  12) ${\bf pure \:} (\lambda x. f x) \rightarrow_{\eta} {\bf pure \:} f$;

  13) ${\bf pure \:} (\langle \pi_1 P, \pi_2 P \rangle) \rightarrow_{\eta} {\bf pure \:} P$;

  14) ${\bf pure \:} ({\bf let \: pure \:} x = M {\: \bf in \: } x) \rightarrow_{\eta} {\bf pure \:} M$;

  15) ${\bf pure \:} ({\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} N) \rightarrow_{\eta} { \bf pure \:} ({\bf pure \:} N)$.
  \end{defin}

@@

  1) Let $\Gamma \vdash x : A$, where $\Gamma = \Delta, x : A$, $x \in \mathbb{V}$.

@@

  i) Let $\Gamma \vdash M : A$ and $M \twoheadrightarrow_{\beta} N$, then $\Gamma \vdash N : A$

  ii) Let $\Gamma \vdash M : A$ and $M \twoheadrightarrow_{\eta} N$, then $\Gamma \vdash N : A$
  \end{theorem}

@@

  i) For multistep $\beta$-reduction:

1) Let $\Gamma \vdash {\bf let \: pure \:} \langle \vec{x}, y, \vec{z} \rangle = \langle \vec{M}, {\bf let \: pure \: } \vec{w} = \vec{N} { \: \bf in \: } Q, \vec{P} \rangle { \: \bf in \: } R : {\bf K }B$

By generation we have $\Gamma \vdash \vec{M} : {\bf K}\vec{A_1}$, $\Gamma \vdash {\bf let \: pure \: } \vec{w} = \vec{N} {\: \bf in \:} Q : {\bf K}\vec{A_2}$, $\Gamma \vdash \vec{P} : {\bf K}\vec{A_3}$
and $\vec{x} : \vec{A_1}, y : A_2, \vec{z} : \vec{A_3} \vdash R : B$.

@@

Hence $\Gamma \vdash {\bf pure \: } M_i : { \bf K } A_i$ by ${\bf K}_I$.

4) Let $\Gamma \vdash {\bf let \: pure \:} (\langle \vec{x}, y, \vec{z} \rangle = \langle \vec{M}, {\bf let \: pure \: } \vec{w} = \vec{N} {\bf \: in \: } Q, \vec{P} \rangle { \: \bf in \: } R) : {\bf K }^2 B$.

By generation $\Gamma \vdash {\bf let \: pure \:} \langle \vec{x}, y, \vec{z} \rangle = \langle \vec{M}, {\bf let \: pure \: } \vec{w} = \vec{N} {\bf \: in \: } Q, \vec{P} \rangle { \: \bf in \: } R : {\bf K }B$,

hence $\Gamma \vdash {\bf let \: pure \:} \langle \vec{x}, \vec{w}, \vec{z} \rangle = \langle \vec{M}, \vec{N}, \vec{P} \rangle \: { \bf in } \: R [y := Q] : {\bf K} B$ by the first case.

So $\Gamma \vdash {\bf pure \:} ({\bf let \: pure \:} \langle \vec{x}, \vec{w}, \vec{z} \rangle = \langle \vec{M}, \vec{N}, \vec{P} \rangle \: { \bf in } \: R [y := Q]) : {\bf K}^2 B$
by ${\bf K}_I$.

ii) For multistep $\eta$-reduction:

  1) Let $\vdash {\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} N : {\bf K} A$.

  Then by generation $\vdash N : A$, so $\vdash {\bf pure \:} N : {\bf K} A$ by ${\bf K}_I$.

  2) Let $\Gamma \vdash {\bf let \: pure \:} x = M {\: \bf in \: } x : {\bf K} A$.

  By generation $\Gamma \vdash M : {\bf K} A$ and $x : A \vdash x : A$, hence $\Gamma \vdash M : {\bf K} A$.

  3) Let $\Gamma \vdash {\bf pure \:} (\lambda x. f x) : {\bf K}(A \to B)$.

  By generation $\Gamma \vdash \lambda x. f x : A \to B$, so $\Gamma \vdash f : A \to B$, then $\Gamma \vdash {\bf pure \:} f : {\bf K}(A \to B)$ by ${\bf K}_I$.

  4) Let $\Gamma \vdash {\bf pure \:} (\langle \pi_1 P, \pi_2 P \rangle) : {\bf K}(A \times B)$.

  By generation $\Gamma \vdash \langle \pi_1 P, \pi_2 P \rangle : A \times B$, then $\Gamma \vdash P : A \times B$.

  By ${\bf K}_I$, $\Gamma \vdash {\bf pure \:} P : {\bf K}(A \times B)$.

  5) $\Gamma \vdash {\bf pure \:} ({\bf let \: pure \:} x = M {\: \bf in \: } x) : {\bf K}^2 A$.

  Then $\Gamma \vdash M : {\bf K}^2 A$ and $x : {\bf K} A \vdash x : {\bf K} A$, so $\Gamma \vdash M : {\bf K}^2 A$.

  6) Let $\vdash {\bf pure \:} ({\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} N) : {\bf K}^2 A$.

  By generation ${\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} N : {\bf K}A$, so $\vdash N : A$, then $\vdash {\bf pure \:} N : {\bf K}$.

@@

\section{Categorical semantics}

\begin{defin} Lax monoidal functor

  Let $\langle \mathcal{C}, \oplus_1, \mathds{1} \rangle$ and $\langle \mathcal{D}, \oplus_2, \mathds{1}' \rangle$ are monoidal categories.

  A lax monoidal functor $\mathcal{F} : \langle \mathcal{C}, \oplus_1, \mathds{1} \rangle \to \langle \mathcal{D}, \oplus_2, \mathds{1}' \rangle$ is a functor
  $\mathcal{F} : \mathcal{C} \to \mathcal{D}$ with additional natural transformations:

  1) $u : \mathds{1}' \to \mathcal{F}\mathds{1}$;

  2) $\ast_{A, B} \mathcal{F}A \otimes_2 \mathcal{F}B \to \mathcal{F}(A \otimes_1 B)$

\begin{defin} Applicative functor

  An applicative functor is a triple $\langle \mathcal{C}, \mathcal{K}, \eta \rangle$,
where $\mathcal{C}$ is a symmetric monoidal category, $\mathcal{K}$ is a lax monoidal endofunctor and $\eta$ is a natural transformation, such that:

1) $u = \eta_{\mathds{1}}$;

2) $\ast_{A,B} \circ (\eta_A \otimes \eta_B) = \eta_{A \otimes B}$;

3) Weak commutativity condition holds:

\xymatrix{
A \otimes \mathcal{K}B & \mathcal{K}A \otimes \mathcal{K}B & \mathcal{K}(A \otimes B) \\
\mathcal{K}B \otimes A & \mathcal{K}B \otimes \mathcal{K}A & \mathcal{K}(B \otimes A)
}
\end{defin}

By default we will consider an arbitrary closed functor on some cartersian closed category, which is the special case of an applicative functor.

\subsection{Soundness}

\begin{defin} Semantical translation from $\lambda_{\textbf{K}}$ to CCC with applicative functor $\mathcal{K}$:

1) Interpretation for types:

$[\![A]\!] := \hat{A}, A \in \mathbb{T}$;

$[\![A \to B]\!] := [\![A]\!] \to [\![B]\!]$;

$[\![A \times B]\!] := [\![A]\!] \times [\![B]\!]$.

2) Interpretation for modal types: $[\![\textbf{K}A]\!] = \mathcal{K}[\![A]\!]$;

3) Interpretaion for contexts:

$[\![\Gamma = \{ x_1 : A_1, ..., x_n : A_n\}]\!] := [\![\Gamma]\!] = [\![A_1]\!] \times ... \times [\![A_n]\!]$;

4) Interpretation for typing assignment: $[\![\Gamma \vdash M : A]\!] := [\![M]\!] : [\![\Gamma]\!] \to
[\![A]\!]$.

5) Interpretation for typing rules:

\begin{prooftree}
\AxiomC{$ $}
\UnaryInfC{$[\![\Gamma, x : A \vdash x : A]\!] = \pi_2 : [\![\Gamma]\!] \times [\![A]\!] \rightarrow
[\![A]\!]$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$[\![\Gamma, x : A \vdash M : B]\!] = f : [\![\Gamma]\!] \times [\![A]\!] \rightarrow [\![B]\!]$}
\UnaryInfC{$[\![\Gamma \vdash (\lambda x. M) : A \to B]\!] = \Lambda(f) : [\![\Gamma]\!]
\rightarrow[\![B]\!]^{[\![A]\!]}$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$[\![\Gamma \vdash M : A \to B]\!] = [\![M]\!] : [\![\Gamma]\!] \rightarrow [\![B]\!]^{[\![A]\!]}$}
\AxiomC{$[\![\Gamma \vdash N : A]\!] = [\![N]\!] : [\![\Gamma]\!] \rightarrow [\![A]\!]$}
\BinaryInfC{$[\![\Gamma \vdash (M N) : B]\!] = [\![\Gamma]\!] \xrightarrow{\langle [\![M]\!], [\![N]\!]
\rangle} [\![B]\!]^{[\![A]\!]} \times [\![A]\!] \xrightarrow{\epsilon} [\![B]\!] $}
\end{prooftree}

\begin{prooftree}
\AxiomC{$[\![\Gamma \vdash M : A ]\!] = f : [\![\Gamma]\!] \rightarrow [\![A]\!]$}
\AxiomC{$[\![\Gamma \vdash N : B ]\!] = g : [\![\Gamma]\!] \rightarrow [\![B]\!]$}
\BinaryInfC{$[\![\Gamma \vdash (M, N) : A \times B]\!] = \langle f, g \rangle : [\![\Gamma]\!] \rightarrow
[\![A]\!] \times [\![B]\!]$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$[\![\Gamma \vdash p : A_1 \times A_2]\!] = f : [\![\Gamma]\!] \rightarrow [\![A_1]\!] \times
[\![A_2]\!]$}
\RightLabel{$i \in \{1,2\}$}
\UnaryInfC{$[\![\Gamma \vdash \pi_i p : A_i]\!] = [\![\Gamma]\!] \xrightarrow{f} [\![A_1]\!] \times
[\![A_2]\!] \xrightarrow{\pi_i} [\![A_i]\!]$}
\end{prooftree}

\begin{prooftree}
\AxiomC{$[\![\Gamma \vdash M : A]\!] = [\![M]\!] : [\![\Gamma]\!] \rightarrow [\![A]\!]$}
\UnaryInfC{$[\![\Gamma \vdash {\bf pure \:} M : \textbf{K}A]\!] := [\![\Gamma]\!] \xrightarrow{[\![M]\!]}
[\![A]\!] \xrightarrow{\eta_{[\![A]\!]}} \mathcal{K}[\![A]\!]$}
\end{prooftree}

  \begin{prooftree}
    \AxiomC{$[\![\Gamma \vdash \vec{M} : {\bf K} \vec{A}]\!] = \langle [\![M_1]\!],\dots, [\![M_n]\!] \rangle : [\![\Gamma]\!] \rightarrow \prod \limits_{i=i}^n \mathcal{K}[\![A_i]\!]$}
    \AxiomC{$[\![\vec{x} : \vec{A} \vdash N : B]\!] = [\![N]\!] : \prod \limits_{i=i}^n [\![A_i]\!] \rightarrow [\![B]\!]$}
    \BinaryInfC{$[\![\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \: } M : {\bf K} B]\!] = \mathcal{K}([\![N]\!]) \circ \ast_{[\![A_1]\!],\dots,[\![A_n]\!]} \circ \langle [\![M_1]\!],\dots, [\![M_n]\!] \rangle : [\![\Gamma]\!] \rightarrow \mathcal{K}[\![B]\!]$}
  \end{prooftree}

\begin{defin} Simultaneous substitution

Let $\Gamma = \{ x_1 : A_1, ..., x_n : A_n \}$, $\Gamma \vdash M : A$ and for all $i \in \{ 1,..., n \}$,
$\Gamma \vdash M_i : A_i$.

We define simultaneous substitution $M [ \vec{x} := \vec{M}]$ recursively by:

1) $x_i [ \vec{x} := \vec{M}] := M_i $;

2) $(\lambda x. M) [ \vec{x} := \vec{M}] := \lambda x. (M [ \vec{x} := \vec{M}])$;

3) $(M N) [ \vec{x} := \vec{M}] = (M [ \vec{x} := \vec{M}]) (N [ \vec{x} := \vec{M}])$;

4) $\langle M, N \rangle = \langle (M [ \vec{x} := \vec{M}]), (N [ \vec{x} := \vec{M}])\rangle$;

5) $(\pi_i P) [ \vec{x} := \vec{M}] = \pi_i (P [ \vec{x} := \vec{M}])$;

6) $(\text{pure } M) [ \vec{x} := \vec{M}] = \text{pure } (M [ \vec{x} := \vec{M}])$;

\begin{lemma}
$ $

$[\![M [x_1 := M_1,\dots, x_n := M_n]]\!] = [\![M]\!] \circ \langle [\![M_1]\!], \dots, [\![M_n]\!] \rangle$.

1) $[\![(\text{pure } M) [ \vec{x} := \vec{M}]]\!] = [\![\text{pure } M]\!] \circ \langle [\![M_1]\!], \dots,
[\![M_n]\!] \rangle$.

$\begin{array}{lll}
& [\![(\text{pure } M) [ \vec{x} := \vec{M}]]\!] = [\![\text{pure } (M [ \vec{x} := \vec{M}])]\!] &
\text{Substitution definition} \\
& \quad\quad\quad\quad\quad\quad\quad = p \circ [\![(M [ \vec{x} := \vec{M}])]\!]&
\text{Translation for pure}\\
& \quad\quad\quad\quad\quad\quad\quad = p \circ [\![M]\!] \circ \langle [\![M_1]\!], \dots,
[\![M_n]\!] \rangle & \text{Induction hypothesis} \\
& \quad\quad\quad\quad\quad\quad\quad = (p \circ [\![M]\!]) \circ \langle [\![M_1]\!], \dots,
[\![M_n]\!] \rangle & \text{Associativity of composition}\\
& \quad\quad\quad\quad\quad\quad\quad = [\![\text{pure } M]\!] \circ \langle [\![M_1]\!], \dots,
[\![M_n]\!] \rangle& \text{Translation for pure}\\
\end{array}$

\begin{theorem} Soundness

  Let $\Gamma \vdash M : A$ and $M =_{\beta\eta} N$, then $[\![\Gamma \vdash M : A]\!] = [\![\Gamma \vdash N : A]\!]$
\end{theorem}
